Probing heavy ion collisions using quark and gluon jet substructure

### 2018-03-07 qg_ML.tex
updated the text - moved around the lund diagrams

Jet substructure measurements provide concrete and consistent physics information of the modification of jets in a heavy ion environment. From the measurements of the jet shape \cite{Ellis:1992qq,Chatrchyan:2013kwa,Khachatryan:2016tfj,Seymour:1997kj,Li:2011hy,Li:2012bw,Vitev:2008rz,Vitev:2009rd} and jet fragmentation function \cite{Procura:2009vm,Chatrchyan:2012gw,Aad:2014wha,Chatrchyan:2014ava,Aaboud:2017bzv,Sirunyan:2018qec}, the community has recently established that both transverse and longitudinal momentum distributions inside jets are significantly modified. Since jet substructures depend strongly on the partonic origin of jets \cite{Gallicchio:2011xq,Gallicchio:2012ez,Chien:2012ur,Chien:2015ctp,Chien:2014nsa}, a change in the fraction of quark-initiated jets and gluon-initiated jets at the LHC may significantly contribute to substructure modifications observed in experiments. It has recently been realized that an increase of the quark jet fraction in heavy ion collisions can constitute the narrowing of jet cores as suggested in various inclusive jet substructure measurements~\cite{Chien:2015hda,Spousta:2015fca}. With jet samples consisting of different quark and gluon jet fractions, for example comparing inclusive jet and photon-tagged jet measurements, one can disentangle the effect on jet substructure modifications due to the changes in quark and gluon jet fractions caused by jet-medium interactions. This then allows one to focus on studying how quark and gluon jets are differently modified and use them as independent jet quenching probes. Note that the above jet modification picture implies that, with the high-purity quark jets in photon-tagged jet samples, the narrowing effect of jet cores should decrease. On the other hand, certain jet quenching models predict a universal narrowing of jet cores \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm,Casalderrey-Solana:2016jvj,Brewer:2017fqy} and emphasize contributions from medium responses \cite{Tachibana:2017syd}. Selecting purer samples of quark and gluon jets will provide the opportunity for directly studying how the medium responds to different hard probes.

@@

Jet quenching studies of the past fifteen years have made compelling, qualitative strides by comparing jets in proton-proton and heavy ion collisions. In this paper, we systematically study the tagging of quark and gluon jets in both proton-proton and lead-lead collisions making full use of their substructure information. This indirect, complimentary approach studies the  differences in the modification of quark v.s. gluon jets thus disentangling modifications of their common features. By first comparing probes in the same collision system, it can also help reduce the systematic uncertainty in experimental measurements. We will demonstrate our general method using Monte Carlo simulations and take \jw as a concrete example \cite{Zapp:2013zya,KunnawalkamElayavalli:2016ttl}. However, the method is easily applicable to other heavy ion jet simulations \cite{Armesto:2009fj,Casalderrey-Solana:2016jvj,Cao:2017zih} and experimental data as well.

We will use three classes of methods and gain insights from comparing their performances. We combine in a multivariate analysis, five physics-motivated and representative jet substructure variables which are effective in quark gluon discrimination \cite{Gallicchio:2012ez}. They include the jet mass, two radial moments including the girth, the $p_T^D$ and the pixel multiplicity. These variables are combined and studied in a multi-layer perceptron (MLP) with two hidden layers to extract correlations among the variables. Next, we employ a modern 2-dimensional image recognition technique involving a deep convolutional neural network (DCNN). The network is trained on discretized images of quark jets and gluon jets on pseudo rapidity-azimuthal angle $(\eta,\phi)$ plane which forms a fixed-dimensional representation of the jet energy distribution. Each jet image undergoes standard pre-processing as described in \cite{deOliveira:2015xxd}. Moreover, in order to systematically extract the complete jet substructure information, we apply the newly developed telescoping deconstruction (TD) method. At each order $N$, TD deconstructs a jet into its fragmentation basis with regard to $N$ axes that are defined using the winner-take-all (WTA) recombination scheme \cite{Bertolini:2013iqa}. The corresponding subjet expansion can be truncated at any finite order thus allowing us to order-by-order examine the information carried by subleading subjets. These three methods are used in discriminating quark and gluon jets, as well as discriminating jets in proton-proton and heavy ion collisions.

Due of the huge underlying event background in heavy ion collisions, jets with smaller radii are more reliably reconstructed and theoretically well understood \cite{Dasgupta:2014yra,Chien:2015cka,Becher:2015hka,Kang:2016mcy}. In order to minimize the background contamination within the experimentally allowed angular resolution, we study jets reconstructed using the anti-$k_t$ algorithm \cite{Cacciari:2008gp} with $R=0.4$. Within the catchment area of such jets, the significant background contamination needs to be subtracted before meaningful jet observable measurements can be made. In order to study multiple jet substructure observables and their correlations simultaneously, it necessarily requires a constituent-level, observable-independent background subtraction method \cite{Soyez:2012hv,Cacciari:2014gra,Berta:2014eza,Bertolini:2014bba,Komiske:2017ubm}. Furthermore, various jet grooming techniques \cite{Ellis:2009su,Ellis:2009me,Krohn:2009th,Dasgupta:2013ihk,Larkoski:2014wba} systematically remove soft radiation and highlights contributions from hard, collinear particles to jet observables. We will draw connections from telescoping deconstruction to the soft-drop groomed momentum sharing $z_g$ and groomed jet radius $r_g$ variables \cite{Larkoski:2014wba} and discuss qualitatively new jet substructure observables within the context of the telescoping deconstruction framework. We will also use Lund diagrams \cite{Andersson1989,Salam:2016yht} to represent and showcase significant regions of phase space in the context of the soft drop grooming procedure.

The rest of the paper is organized as follows. In Sec.~\ref{sec:sample} we describe the jet samples we generate using the \jw simulation. In Sec.~\ref{sec:qg} we give details about the multivariate analysis, the jet image, and the telescoping deconstruction methods, as well as the corresponding jet substructure distributions. In Sec.~\ref{sec:results} we compare the performances of the methods and discuss the physical interpretations. In Sec.~\ref{sec:lund} we use Lund diagrams to illustrate the regions of phase space in jet formation separated by the soft-drop procedure. We conclude in Sec.~\ref{sec:conc} and raise awareness of such methods by giving an outlook to future studies.

@@

To use quark and gluon jets as independent probes of heavy ion collisions, we need to first define and identify the respective probes. In this paper, we define jet flavor based on the parton from hard scattering that initiates the parton shower evolution. We use the \jw Monte Carlo simulation, a perturbative jet evolution framework to generate proton-proton and central (0-20\%) lead-lead collisions at $\sqrt{s}=2.76$ TeV using the standard setup \cite{Zapp:2013zya}. We use the prompt photon production channels \cite{KunnawalkamElayavalli:2016ttl} $q +\gamma$ and $g +\gamma$ to generate quark and gluon jet samples, respectively, as an attempt to prepare two sets of jets with distinct quark and gluon jet fractions. It is possible to construct jet samples by performing hard process and kinematical selections (such as dijet v.s. $Z$+jet) to enhance either the gluon or the quark jet fraction. On the other hand, weak and unsupervised learning algorithms \cite{Dery:2017fap, Cohen:2017exh, Metodiev:2017vrx,Komiske:2018oaa,Metodiev:2018ftz} can also be employed in the near future on such samples in order to treat both data and simulations on equal footing.

In each simulated event, jets are reconstructed using the anti-$k_t$ algorithm~\cite{Cacciari:2008gp} as implemented in \textsc{FastJet}~\cite{Cacciari:2011ma} with jet radius $R = 0.4$. We impose the following kinematic cuts,
\begin{equation}
    p^{\gamma}_T > 100~{\rm GeV}, ~~~ |\eta^{\rm \gamma}|<1.5, ~~~ p^{\rm jet}_T > 50~{\rm GeV}, ~~~ |\eta^{\rm jet}|<1.5, ~~~\Delta \phi_{\gamma, {\rm jet}} > 2\pi/3\;,
\end{equation}
where $p_T$, $\eta$ and $\phi$ are the transverse momentum, pseudo rapidity and azimuthal angle, respectively, to select high $p_T$ jets recoiling against the prompt photon. Jet quenching in central PbPb collisions in \jw is calculated based on a medium model consisting of thermal scattering centers undergoing Bjorken expansion. The medium formation time is set to $\tau_\text{i}=0.6 $ fm along with an initial temperature $T_\text{i}=485$ MeV based on hydrodynamic calculations~\cite{Shen:2012vn,Shen:2014vra}. We use the \textsc{CTEQ6LL} \cite{Pumplin:2002vw} and the \textsc{EPS09}~\cite{Eskola:2009uj} parton distribution functions in \textsc{LHAPDF-5}~\cite{Whalley:2005nh}. We include recoil partons in our PbPb simulations that originate from elastic and inelastic scattering with thermal scattering centers due to interactions with propagating hard scattered partons. It was recently shown that the recoil contributions is important in \jw to accurately describe several qualitative features as seen in jet substructure modifications measured at the LHC \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm}. In order to remove the thermal component of the recoil partons, we adopt the background subtraction procedure introduced in \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm}. The \jw simulation produces HepMC~\cite{Dobbs:2001ck} files which are processed using a \textsc{Rivet} analysis framework~\cite{Buckley:2010ar}.

To emulate finite detector effects on the jet angular resolution, we discretize jets using $\eta-\phi$ grids of size $0.08 \times 0.08$ respecting the current angular resolution of hadronic calorimeter at the LHC. To study how such discretization as well as background subtraction affect the analysis, we use the following three different sets of jets in heavy ion collisions, all with the medium recoil contributions,
\begin{itemize}
	\item neither discretization nor background subtraction
	\item with discretization but without background subtraction
	\item with discretization and the GridSub \cite{KunnawalkamElayavalli:2017hxo} background subtraction.
\end{itemize}
For jets in proton-proton collisions, we use the sets of jet samples with and without discretization. In experiments, the fluctuating, underlying event background is subtracted via specific background subtraction techniques. Since \jw does not generate full underlying events in quenched samples, we can not directly apply those background subtraction methods. With the application of the grid subtraction (GridSub) technique, the collection of jets in \jw represent ideally subtracted jets in the experiment. This enables us to focus on and highlight the effects of jet quenching, as opposed to additional, nontrivial smearing effects.

@@

@@

Figure~\ref{fig:jetdistributons_pp_pbpb} shows the distributions of the aforementioned jet observables in pp and PbPb collisions for quark (blue shaded histograms) and gluon jets (orange shaded histograms). Both pp and PbPb jets include the grid discretization. Furthermore, jets in PbPb collisions are subtracted using the GridSub background subtraction method. Note that the jet mass and radial moments are infrared and collinear (IRC) safe observables with different weights on the particle angular distributions. In vacuum, they reflect the fact that gluon jets have larger jet mass and larger radial moments than quark jets due to the larger color charge and Casimir color factor $C_A>C_F$. In the \jw PbPb simulations, both quark and gluon jets have distributions that are modified towards larger values. On the other hand, pixel multiplicity and the $p_T^D$ are IRC unsafe observables. We see a significant increase of the pixel multiplicity and decrease of the values of $p_T^D$ for quark and gluon jets, respectively.

@@

The multivariate analysis discussed previously is a constructive approach to collect and examine the usefulness of jet features that may help distinguish quark jets from gluon jets, or jets in different collision systems. An alternative approach is to use image-recognition techniques which optimize comprehensive multivariate models and identify all possible useful features. The radiation pattern of a jet can be quantified as an image in the $\eta-\phi$ plane with $p_T$ deposited in detector cells, much like a digital camera. The jet image thus provides a fixed-dimensional representation of jet energy distribution. Each jet image is pre-processed \cite{deOliveira:2015xxd} by translating in the $\eta-\phi$ plane so that the highest pixel is centered at the origin. Also, the image is rotated so that the principal component of the energy density lies along the $\eta=0, \phi<0$ direction.

@@

A deep convolutional neural network is implemented in Keras with the TensorFlow \cite{DBLP:journals/corr/AbadiBCCDDDGIIK16} backend. The neural network model consists of three convolution layers of sequentially reducing sizes, each with 8 filters followed by a max-pooling layer. The convolution layers use hyperbolic tangent activation function, and the output layer uses sigmoid activation function. The output layer is preceded by three layers, a dropout layer with a rejection score of 0.20, a dense layer with 20 nodes and an additional dropout layer with a reduced rejection score of 0.10. The additional dropout layer helps filtering less important features from previous layers and thus increases the speed and efficiency of the model training. The model is trained using the Adam optimizer with binary cross entropy loss function.

@@

Beside the overall classification performance using telescoping deconstruction we present in the next section, subjet variables within this framework can reveal many aspects of jet dynamics. In particular, there is rich information contained in subjet topology and subjet substructure. Recently, the CMS~\cite{Sirunyan:2017bsd} and STAR~\cite{Kauder:2017mhg} collaborations measured the groomed momentum sharing observable $z_g$. One reclusters jets using the Cambridge/Aachen (C/A) algorithm \cite{Dokshitzer:1997in,Wobisch:1998wt} and sequentially removes wide angle, soft radiation until the following condition is satisfied by a branching,
\begin{equation}
    z_{\rm cut} < \frac{\min(p_{T_1},p_{T_2})}{p_{T_1}+p_{T_2}} \equiv z_g\;.
\label{SD}
\end{equation}
Here $p_{T_1}$ and $p_{T_2}$ are the transverse momenta of the two branches, and $z_{\rm cut}$ is the soft-drop parameter which is set to $0.1$ in our analysis. Note that the C/A clustering tree is angular-ordered and therefore the above procedure identifies the most wide angle, soft subjet that carries a significant fraction of the jet transverse momentum larger than $z_{\rm cut}$. The sizes of the subjets are dynamically determined and related to the angle $r_g$ between the two soft-drop branches, defined as the groomed jet radius. The observables $z_g$ and $r_g$ thus encode the kinematic information of the two subjets which is in the category of subjet topology. In vacuum, the observable $z_g$ is closely related to the Altarelli-Parisi splitting functions~\cite{Altarelli:1977zs}. The physical meaning of $z_g$ in heavy ion collisions have also been actively investigated \cite{Chien:2016led,Mehtar-Tani:2016aco,Milhano:2017nzm,Chang:2017gkt,Li:2017wwc}.

@@

In addition to the subjet momentum fraction which provides information about the longitudinal momentum distribution, the angular separation between the two subjets encodes information about the momentum distribution transverse to the jet direction. Figure~\ref{fig:comp_rg} shows the distributions of the groomed jet radius $r_g$ (left panels) and the angle $\theta_{\rm T3}$ (right panels) for quark (blue) and gluon (red) jets. %Note that the value of $r_g$ is effectively cutoff by the jet radius $R$, while the value of $\theta_{\rm T3}$ can be larger than $R$.
Both the observables have similar distributions for quark and gluon jets in vacuum as shown in the top panels of Figure~\ref{fig:comp_rg}. However, we see a significant enhancement of the $r_g$ distribution at large angle ($r_g\approx0.3$) near the edge of the quenched jets, while the modification of $\theta_{\rm T3}$ has a similar enhancement but with smaller magnitude (bottom panels). Note the difference in the subjet axis definitions which affects the angles among them: soft-drop uses C/A axes whereas TD uses WTA axes. Also, the subjet radii in TD provide an extra handle on the area of coverage by subjets therefore soft radiation can be probed with various angular resolutions. While these observables have not yet been experimentally measured at the RHIC or the LHC, the observed phenomenon of jet broadening in the \jw Monte Carlo simulations is consistent with the jet shape measurement~\cite{Chatrchyan:2013kwa}.

@@

\begin{figure}[t]
	   \centering
	   \includegraphics[width=0.9\textwidth]{plots/Comp_lambda_new.pdf}
	   \caption{Distributions of TD $\lambda$, which is the ratio between subjet mass and subjet $p_T$ for hard (top) and soft (bottom panels) subjets with quark (blue dashed) and gluon (red dashed) jets in pp (left) and central PbPb (right) collisions in \textsc{Jewel}. }
\label{fig:comp_subjet_m}
\end{figure}

>>>>

Jet substructure measurements provide concrete and consistent physics information of the modification of jets in a heavy ion environment. From the measurements of the jet shape \cite{Ellis:1992qq,Chatrchyan:2013kwa,Khachatryan:2016tfj,Seymour:1997kj,Li:2011hy,Li:2012bw,Vitev:2008rz,Vitev:2009rd} and jet fragmentation function \cite{Procura:2009vm,Chatrchyan:2012gw,Aad:2014wha,Chatrchyan:2014ava,Aaboud:2017bzv,Sirunyan:2018qec}, the community has recently established that both transverse and longitudinal momentum distributions inside jets are significantly modified. Since jet substructures depend strongly on the partonic origin of jets \cite{Gallicchio:2011xq,Gallicchio:2012ez,Chien:2012ur,Chien:2015ctp,Chien:2014nsa}, a change in the fraction of quark-initiated jets and gluon-initiated jets at the LHC may significantly contribute to substructure modifications observed in experiments. It has recently been realized that an increase of the quark jet fraction in heavy ion collisions can contribute to the narrowing of jet cores as suggested in various inclusive jet substructure measurements~\cite{Chien:2015hda,Spousta:2015fca}. With jet samples consisting of different quark and gluon jet fractions, for example comparing inclusive jet and photon-tagged jet measurements, one can disentangle the effect on jet substructure modifications due to the changes in quark and gluon jet fractions caused by jet-medium interactions. This then allows one to focus on studying how quark and gluon jets are differently modified and use them as independent jet quenching probes. Note that the above jet modification picture implies that, with the high-purity quark jets in photon-tagged jet samples, the narrowing effect of jet cores should decrease. On the other hand, certain jet quenching models predict a universal narrowing of jet cores \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm,Casalderrey-Solana:2016jvj,Brewer:2017fqy} and emphasize contributions from medium responses \cite{Tachibana:2017syd}. Selecting purer samples of quark and gluon jets will provide the opportunity for directly studying how the medium responds to different hard probes.

@@

Jet quenching studies of the past sixteen years have made compelling, qualitative strides by comparing jets in proton-proton and heavy ion collisions. In this paper, we systematically study the tagging of quark and gluon jets in both proton-proton and lead-lead collisions making full use of their substructure information. This indirect, complimentary approach studies the  differences in the modification of quark v.s. gluon jets, thus disentangling modifications of their common features in a multivariate analysis. By first comparing probes in the same collision system, it can also help reduce the systematic uncertainty in experimental measurements. We will demonstrate our general method using Monte Carlo simulations and take \jw as a concrete example \cite{Zapp:2013zya,KunnawalkamElayavalli:2016ttl}. However, the method is easily applicable to other heavy ion jet simulations \cite{Armesto:2009fj,Casalderrey-Solana:2016jvj,Cao:2017zih} and experimental data as well.

We will use three classes of methods and gain insights from comparing their performances. We combine in a multivariate analysis, five physics-motivated and representative jet substructure variables which are effective in quark gluon discrimination \cite{Gallicchio:2012ez}. They include the jet mass, two radial moments including the girth, the $p_T^D$ and the pixel multiplicity. These variables are combined and studied in a multi-layer perceptron (MLP) with two hidden layers to extract correlations among the variables. Next, we employ a modern 2-dimensional image recognition technique involving a deep convolutional neural network (DCNN). The network is trained on discretized images of quark jets and gluon jets on pseudo rapidity-azimuthal angle $(\eta,\phi)$ plane which forms a fixed-dimensional representation of the jet energy distribution. Each jet image undergoes standard pre-processing as described in \cite{deOliveira:2015xxd}. Finally, in order to systematically extract the complete jet substructure information, we apply the newly developed telescoping deconstruction (TD) method. At each order $N$, TD deconstructs a jet into its fragmentation basis with regard to $N$ axes that are defined using the winner-take-all (WTA) recombination scheme \cite{Bertolini:2013iqa}. The corresponding subjet expansion can be truncated at any finite order thus allowing us to order-by-order examine the information carried by subleading subjets. These three methods are used in discriminating quark and gluon jets, as well as discriminating jets in proton-proton and heavy ion collisions.

Due of the huge underlying event background in heavy ion collisions, jets with smaller radii are more reliably reconstructed and theoretically well understood \cite{Dasgupta:2014yra,Chien:2015cka,Becher:2015hka,Kang:2016mcy}. In order to minimize the background contamination within the experimentally allowed angular resolution, we study jets reconstructed using the anti-$k_t$ algorithm \cite{Cacciari:2008gp} with $R=0.4$. Within the catchment area of such jets, the significant background contamination needs to be subtracted before meaningful jet observable measurements can be made. In order to study multiple jet substructure observables and their correlations simultaneously, it necessarily requires a constituent-level, observable-independent background subtraction method \cite{Soyez:2012hv,Cacciari:2014gra,Berta:2014eza,Bertolini:2014bba,Komiske:2017ubm}. Furthermore, various jet grooming techniques \cite{Ellis:2009su,Ellis:2009me,Krohn:2009th,Dasgupta:2013ihk,Larkoski:2014wba} systematically remove soft radiation and highlight contributions from hard, collinear particles to jet observables. We will draw connections from telescoping deconstruction to the soft-drop groomed momentum sharing $z_g$ and groomed jet radius $r_g$ variables \cite{Larkoski:2014wba} and discuss qualitatively new jet substructure observables within the context of the telescoping deconstruction framework. During the course of the paper, we use Lund diagrams \cite{Andersson1989,Salam:2016yht} to represent and showcase significant regions of phase space in the context of jet quenching in \jw and with regards to the soft drop grooming procedure.

The rest of the paper is organized as follows. In Sec.~\ref{sec:sample} we describe the jet samples we generate using the \jw simulation. In Sec.~\ref{sec:qg} we give details about the multivariate analysis, the jet image, and the telescoping deconstruction methods, as well as the corresponding jet substructure distributions. In Sec.~\ref{sec:results} we compare the performances of the methods and discuss the physical interpretations. We conclude in Sec.~\ref{sec:conc} and raise awareness of such methods by giving an outlook to future studies.

@@

To use quark and gluon jets as independent probes of heavy ion collisions, we need to first define and identify the respective probes. In this paper, we define jet flavor based on the parton from hard scattering that initiates the parton shower evolution. We use the \jw Monte Carlo simulation, a perturbative jet evolution framework to generate proton-proton and central (0-20\%) lead-lead collisions at $\sqrt{s}=2.76$ TeV using the standard setup \cite{Zapp:2013zya}. We use the prompt photon production channels \cite{KunnawalkamElayavalli:2016ttl} $q +\gamma$ and $g +\gamma$ to generate quark and gluon jet samples, respectively, as an attempt to prepare two sets of jets with distinct quark and gluon jet fractions. It is possible to construct jet samples by performing hard process and kinematical selections (such as dijet v.s. $Z$+jet~\footnote{It is important to note that dijets in heavy ion events are both expected to be quenched by the medium, which can in turn lead to artificially altered quark and gluon fractions.}) to enhance either the gluon or the quark jet fraction. On the other hand, unsupervised or weakly supervised learning algorithms \cite{Dery:2017fap, Cohen:2017exh, Metodiev:2017vrx,Komiske:2018oaa,Metodiev:2018ftz} can also be employed in the near future on such samples in order to treat both data and simulations on equal footing.

In each simulated event, jets are reconstructed using the anti-$k_t$ algorithm~\cite{Cacciari:2008gp} as implemented in \textsc{FastJet}~\cite{Cacciari:2011ma} with jet radius $R = 0.4$. We impose the following kinematic cuts,
\begin{equation}
    p^{\gamma}_T > 100~{\rm GeV}, ~~~ |\eta^{\rm \gamma}|<1.5, ~~~ p^{\rm jet}_T > 50~{\rm GeV}, ~~~ |\eta^{\rm jet}|<1.5, ~~~\Delta \phi_{\gamma, {\rm jet}} > 2\pi/3\;,
\end{equation}
where $p_T$, $\eta$ and $\phi$ are the transverse momentum, pseudo rapidity and azimuthal angle, respectively, to select high $p_T$ jets recoiling against the prompt photon. Jet quenching in central PbPb collisions in \jw is calculated based on a medium model consisting of thermal scattering centers undergoing Bjorken expansion. The medium formation time is set to $\tau_\text{i}=0.6 $ fm along with an initial temperature $T_\text{i}=485$ MeV based on hydrodynamic calculations~\cite{Shen:2012vn,Shen:2014vra}. We use the \textsc{CTEQ6LL} \cite{Pumplin:2002vw} and the \textsc{EPS09}~\cite{Eskola:2009uj} parton distribution functions in \textsc{LHAPDF-5}~\cite{Whalley:2005nh}. We include recoil partons in our PbPb simulations that originate from elastic and inelastic scattering with thermal scattering centers due to interactions with propagating hard scattered partons. It was recently shown that the recoil contributions are important in \jw to accurately describe several qualitative features as seen in jet substructure modifications measured at the LHC \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm}. In order to remove the thermal component of the recoil partons, we adopt the background subtraction procedure introduced in \cite{KunnawalkamElayavalli:2017hxo,Milhano:2017nzm}. The \jw simulation produces HepMC~\cite{Dobbs:2001ck} files which are processed using a \textsc{Rivet} analysis framework~\cite{Buckley:2010ar}.

To emulate finite detector effects on the jet angular resolution, we discretize jets using $\eta-\phi$ grids of size $0.08 \times 0.08$ respecting the current angular resolution of the CMS experiment's inherent hadronic calorimeter resolution at the LHC. To study how such discretization as well as background subtraction affect the analysis, we use the following three different sets of jets in heavy ion collisions, all with the medium recoil contributions,
\begin{itemize}
	\item neither discretization nor background subtraction
	\item with discretization but without background subtraction
	\item with discretization and the GridSub \cite{KunnawalkamElayavalli:2017hxo} background subtraction.
\end{itemize}
For jets in proton-proton collisions, we use the sets of jet samples with and without discretization. In experiments, the fluctuating, underlying event background is subtracted via specific background subtraction techniques. Since \jw does not generate full underlying events in quenched samples, we can not directly apply those background subtraction methods. With the application of the grid subtraction (GridSub) technique, the collection of jets in \jw represent ideally subtracted jets in the experiment. This enables us to focus on and highlight the effects of jet quenching, as opposed to additional, nontrivial smearing effects.

@@

Every jet observable is defined as a function of the constituent momenta and angles from jet axes so as to probe the entire jet substructure with a single number. One can choose a set of jet observables to be a representation of jets and we discus several approaches in the coming sections that faithfully encode the jet information. We use the Lund diagram~\cite{Andersson1989} associated with the Cambridge/Aachen clustering tree structure as diagrammatic jet representation through which one can comprehend the quenching mechanism in \textsc{Jewel}. The Lund diagram directly records the branching kinematics along the hard branches in the C/A clustering tree in terms of longitudinal momentum fraction $z$ of the soft branch and the angular separation $\Delta R$ between the two branches. The Lund diagram is then a two-dimensional histogram of variables $\ln (R/\Delta R)$ and $\ln z(\Delta R/R)$ where $R$ is the jet radius used in jet reconstruction. The variables are chosen such that the Lund diagram is uniform in the fixed strong coupling constant limit for jets in pp collisions due to the infrared structure of QCD and as such, they have been shown to be useful in distinguishing between QCD and boosted heavy particle jets \cite{Salam:2016yht}.

Given a jet reconstructed using the anti-$k_{t}$ algorithm with $R = 0.4$, the procedure of creating a Lund diagram is as follows:
\begin{itemize}
	\item Recluster the jet using the C/A algorithm
	\item Start from the root of the tree and move along the hard branches. At each branching, construct the momentum fraction of the soft branch $z=\min(p_{T_1},p_{T_2})/(p_{T_1}+p_{T_2})$ and the angular distance between the two branches $\Delta R = \sqrt{\Delta \eta^2_{1, 2}+\Delta \phi^2_{1, 2}}$
	\item Fill in the Lund diagram for each branching
	\item Continue along the subsequent hard branches until reaching only one particle
\end{itemize}

In Lund diagrams. wide angle, soft radiation populates close to the $y$-axis, while collinear, hard radiation is along the diagonal (with the negative slope). Figure:~\ref{fig:Lund_full} shows the Lund diagrams for quark (left) and gluon (right) jets in pp (top) and PbPb (bottom) collisions in \textsc{Jewel}. We see a significantly enhanced region close to the $y$-axis in PbPb collisions that corresponds to increased wide angle, soft radiation in quenched jets. Comparing quark to gluon jets, we see that gluon jets (right panel) have a larger fraction of soft radiation. On the other hand, the hard, collinear radiation region of the Lund diagram is shown to reduce in its significance for quenched jets. This modification manifests in modified jet cores in PbPb collisions where subleading subjets tend to have wider angles and softer momenta. These are the high level features we hope to extract with jet representations in order to effectively tag quark jets vs gluon jets in both vacuum and in the medium.

@@

Figure~\ref{fig:jetdistributons_pp_pbpb} shows the distributions of the aforementioned jet observables in pp and PbPb collisions for quark (blue shaded histograms) and gluon jets (orange shaded histograms). Both pp and PbPb jets include the grid discretization. Furthermore, jets in PbPb collisions are subtracted using the GridSub background subtraction method. Note that the jet mass and radial moments are infrared and collinear (IRC) safe observables with different weights on the particle angular distributions. In vacuum, they reflect the fact that gluon jets have larger jet mass and larger radial moments than quark jets due to the larger color charge and Casimir color factor $C_A>C_F$. In the \jw PbPb simulations, both quark and gluon jets have distributions that are modified towards larger values. On the other hand, we see a significant increase of the pixel multiplicity and decrease of the values of $p_T^D$ for quark and gluon jets, respectively, with the caveat that pixel multiplicity is unreliable in a heavy ion environment and the $p_T^D$ is an IRC unsafe observable.

@@

The multivariate analysis discussed previously is a constructive approach to collect and examine the usefulness of jet features that may help distinguish quark jets from gluon jets, or jets in different collision systems. An alternative approach is to use image-recognition techniques which optimize comprehensive multivariate models and identify all possible useful features. The radiation pattern of a jet can be quantified as an image in the $\eta-\phi$ plane with $p_T$ deposited in detector cells, much like a digital camera. The jet image thus provides a fixed-dimensional representation of jet energy distribution. Each jet image is pre-processed \cite{deOliveira:2015xxd} by translating in the $\eta-\phi$ plane so that the most active pixel is centered at the origin. Also, the image is rotated so that the principal component of the energy density lies along the $\eta=0, \phi<0$ direction.

@@

A deep convolutional neural network is implemented in Keras with the TensorFlow \cite{DBLP:journals/corr/AbadiBCCDDDGIIK16} backend. The neural network model consists of three convolution layers of sequentially reducing sizes, each with 8 filters followed by a max-pooling layer. The convolution layers use a hyperbolic tangent activation function, and the output layer uses a sigmoid activation function. The output layer is preceded by three layers, a dropout layer with a rejection score of 0.20, a dense layer with 20 nodes and an additional dropout layer with a reduced rejection score of 0.10. The additional dropout layer helps filtering less important features from previous layers and thus increases the speed and efficiency of the model training. The model is trained using the Adam optimizer with binary cross entropy loss function.

@@

Beside the overall classification performance using telescoping deconstruction we present in the next section, subjet variables within this framework can reveal many aspects of jet dynamics. In particular, there is rich information contained in subjet topology and subjet substructure. Recently, the CMS~\cite{Sirunyan:2017bsd} and STAR~\cite{Kauder:2017mhg} collaborations measured the groomed momentum sharing observable $z_g$. Jets are reclustered using the Cambridge/Aachen (C/A) algorithm \cite{Dokshitzer:1997in,Wobisch:1998wt} and wide angle, soft radiation are sequentially removed until the following condition is satisfied by a branching,
\begin{equation}
    z_{\rm cut} < \frac{\min(p_{T_1},p_{T_2})}{p_{T_1}+p_{T_2}} \equiv z_g\;.
\label{SD}
\end{equation}
Here $p_{T_1}$ and $p_{T_2}$ are the transverse momenta of the two branches, and $z_{\rm cut}$ is the soft-drop parameter which is set to $0.1$ in our analysis. Note that the C/A clustering tree is angular-ordered and therefore the above procedure identifies the most wide angle, soft subjet that carries a significant fraction of the jet transverse momentum larger than $z_{\rm cut}$. The sizes of the subjets are dynamically determined and related to the angle $r_g$ between the two soft-drop branches, defined as the groomed jet radius. The observables $z_g$ and $r_g$ thus encode the kinematic information of the two subjets which is in the category of subjet topology. In vacuum, the observable $z_g$ is closely related to the Altarelli-Parisi splitting functions~\cite{Altarelli:1977zs}. The physical meaning of $z_g$ in heavy ion collisions have also been actively investigated \cite{Chien:2016led,Mehtar-Tani:2016aco,Milhano:2017nzm,Chang:2017gkt,Li:2017wwc}.

@@

In addition to the subjet momentum fraction which provides information about the longitudinal momentum distribution, the angular separation between the two subjets encodes information about the momentum distribution transverse to the jet direction. Figure~\ref{fig:comp_rg} shows the distributions of the groomed jet radius $r_g$ (left panels) and the angle $\theta_{\rm T3}$ (right panels) for quark (blue) and gluon (red) jets. %Note that the value of $r_g$ is effectively cutoff by the jet radius $R$, while the value of $\theta_{\rm T3}$ can be larger than $R$.
Both the observables have similar distributions for quark and gluon jets in vacuum as shown in the top panels of Figure~\ref{fig:comp_rg}. However, we see a significant enhancement of the $r_g$ distribution at large angle ($r_g\approx0.3$) near the edge of the quenched jets, while the modification of $\theta_{\rm T3}$ has a similar enhancement but with smaller magnitude (bottom panels). Note the difference in the subjet axis definitions which affects the angles among them: soft-drop uses C/A axes whereas TD uses WTA axes. Also, the subjet radii in TD provide an extra handle on the area of coverage by subjets therefore soft radiation can be probed with various angular resolutions. While these observables have not yet been experimentally measured at the RHIC or the LHC, the observed phenomenon of jet broadening in the \jw simulations is consistent with the jet shape measurement~\cite{Chatrchyan:2013kwa}.

@@

In vacuum, we see that all the $\{z_g,r_g\}$ and $\{z_{\rm T3},\theta_{\rm T3}\}$ distributions are relatively insensitive to the partonic origin of jets because the color factors cancel in the normalization of the distributions. They encode information about the leading-order subjet topology which is determined by the QCD splitting kinematics.

\begin{figure}[t]
	   \centering
	   \includegraphics[width=0.9\textwidth]{plots/Comp_lambda_new.pdf}
	   \caption{Distributions of TD $\lambda$, which is the ratio between subjet mass and subjet $p_T$ for hard (left) and soft (right panels) subjets with quark (blue dashed) and gluon (red dashed) jets in pp (top) and central PbPb (bottom) collisions in \textsc{Jewel}. }
\label{fig:comp_subjet_m}
\end{figure}

To go beyond to higher-order jet features, we also study the subjet substructure, for example subjet masses. These observables disentangle the subjet kinematic information and are more directly sensitive to the soft radiation around subjets. Figure~\ref{fig:comp_subjet_m} shows the distributions of $\lambda\equiv m/p_T$, which is the ratio between subjet mass and subjet $p_T$, of hard subjets (left panels) and soft subjets (right panels) for quark (blue dashed) and gluon (red dashed) jets. Here hard or soft refers to subjet momenta with harder meaning the one with higher transverse momentum. We see hints that the $\lambda$ distributions of hard subjets exhibit flavor dependence, while those of soft subjets are similar for quark and gluon jets. This is consistent with the physics picture that for a quark jet the hard subjet is typically a quark-initiated subjet, while the soft subjet is a gluon-initiated subjet through $q\rightarrow qg$ splitting. On the other hand, both the subjets of a gluon jet are mostly gluon-initiated subjets through the $g\rightarrow gg$ splitting. We also see the effect of quenching (right panels in Figure~\ref{fig:comp_subjet_m}), where the hard and soft subjet $\lambda$ distributions get modified significantly, especially for soft subjets which tend towards larger values with an enhanced tail. This suggests that the qualitative feature of the soft event activities in the \jw simulations, persists to affect not only the subjet topology but also the subjet substructure.
